Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-06-28 article.tex
Minor changes in discussion. Updated some citations.

\section{Discussion / Implications of a successful magnitude predictor}
In this work we have introduced the concept of a \textit{magnitude predictor} to assist us in finding dependencies between magnitudes and the seismic history.  By focusing solely on magnitude prediction, assuming known location and timing of the event, we have shown the some of the information about an earthquakes magnitude is contained in the seismic history of a simple hypocenter catalog. Throughout this work we have made significant effort to address possible origins of this information gain. We have ruled out temporal and spatial variability in the completeness magnitude - and artifact originating from the seismic network's limitations. Furthermore, we examine a several benchmark models to make sure the advantage of our model, MAGNET, does not originate from a simple, previously attempted, statistical pattern of earthquake occurrence. Among the benchmarks we examine are standard spatially and temporally varying Gutenberg Richter distributions, as well as other benchmarks encompassing the near past seismic trends, such as a kernel density estimation of the past $n$ events. We have demonstrated that MAGNET's performance exceeds that of every benchmark tested, under any examined conditioning \neri{There is that temporal conditioning of the spatial GR I'm not sure how to treat in this sentence...}. It is then implied that MAGNET extracts previously unknown patterns of embodied in the seismic history that hold information of the magnitude of the upcoming earthquake.
A corollary follows this finding: for some earthquakes, the magnitude might be partially predetermined before the earthquake is initiated. This suggests the existence of a precursory process influencing an earthquake's magnitude. However, the specific nature and seismic patterns associated with this precursory stage remain unclear based on our current evidence. Nonetheless, this finding provides a promising avenue for future research aimed at identifying these precursory signals.
Investigating the properties of such precursory signals may be carried out by interpretability analysis of neural networks \neri{cite}. Such method may reveal what functions of the encoders (see architecture in methods section) have a larger weight in the result of well predicted events. Furthermore, the method of ablation \neri{cite} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.

\section{Possible affects on operational earthquake forecasting and prediction}
Discovery of specific pattern in the seismic history as precursory signals can advance the effort of earthquake prediction a great deal.\neri{cite? what?}. Nevertheless, discovering such signals in enough accuracy to prove useful for operational systems may be a further down the road.
An immediate application of the model presented in this work may be of quick magnitude estimation to be used for early warning systems. Such a system does not require additional data sources of assumptions on the model as the timing and location of the earthquake to be estimated are given by design of the task. 
A second application may incorporating such a magnitude predictor into an earthquake rate forecasting model such as the epidemic type aftershock sequence model (ETAS) \neri{cite}. In ETAS, the rate of events is determined by the product of a set of functions. \neri{formalize?} Origin time of events of the forecasted sequence are then sampled from the rate function, and assigned with a magnitude by sampling from a fitted GR distribution. y replacing the randomly assigned magnitudes of ETAS with MAGNET or a similar magnitude predictor one will improve the predictive power of rate prediction. As we have shown - magnitudes predicted by MAGNET will estimate the true magnitudes in a higher likelihood.

>>>>

\section{Discussion / Implications of a successful magnitude predictor}
In this work we have introduced the concept of a \textit{magnitude predictor} to assist us in finding dependencies between magnitudes and the seismic history.  By focusing solely on magnitude prediction, assuming known location and timing of the event, we have shown the some of the information about an earthquakes magnitude is contained in the seismic history of a simple hypocenter catalog. Throughout this work we have made significant effort to account for information gain that may originate in artifacts rather than true properties of the earthquake physical system. We have ruled out temporal and spatial variability in the completeness magnitude - and artifact originating from the seismic network's limitations. Furthermore, we examine several benchmark models to make sure the advantage of our model, MAGNET, does not originate from a simple, previously attempted, statistical pattern of earthquake occurrence. Among the benchmarks we examine are standard spatially and temporally varying Gutenberg Richter distributions, as well as other benchmarks encompassing the near past seismic trends, such as a kernel density estimation of the past $n$ events. We have demonstrated that MAGNET's performance exceeds that of every benchmark tested, under any examined conditioning \neri{There is that temporal conditioning of the spatial GR I'm not sure how to treat in this sentence...}. It is then implied that MAGNET extracts previously unknown patterns of embodied in the seismic history that hold information of the magnitude of the upcoming earthquake.
A corollary follows this finding: for some earthquakes, the magnitude might be partially predetermined before the earthquake is initiated. This suggests the existence of a precursory process influencing an earthquake's magnitude. However, the specific nature and seismic patterns associated with this precursory stage remain unclear based on our current evidence. Such a precursory stage has been proposed in the past, specifically for large earthquakes, by Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020}. We propose the phenomenon we observe in our work may coincide with the work of Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020} as both examine statistical properties leading to emergence of a specific magnitude in determined space and time coordinates. In order to show such competability between to works further investigation is required. Specifically, we propose to examine the time a length scale upon which MAGNET's prediction are valid, ahead of origin time. This can be done be examining query coordinates around and before the true time-space coordiantes  of the labeled events. This can also be done by constructing a different model trained to estimate the magnitude of an upcoming earthquake ahead of the origin time, examining how its accuracy declines with lead time and distance from the event location. 
Additionally, MAGNET may be used in it's current trained state to shed light on potential precursory signals.
% Nonetheless, this finding provides a promising avenue for future research aimed at identifying these precursory signals. Investigating the properties of such precursory signals may be carried out
By explainability analysis of neural networks \cite{sturmfels_visualizing_2020, zhang_survey_2021, liu_interpretable_2023} one may conclude on the statistical patterns ~MAGNET has uncovered. Such method may reveal what functions of the encoders (see architecture in methods section) have a larger weight in the result of well predicted events. Furthermore, the method of ablation \cite{meyes_ablation_2019, sturmfels_visualizing_2020} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.

\section{Possible affects on operational earthquake forecasting and prediction}
Discovery of specific pattern in the seismic history as precursory signals can advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}. Nevertheless, discovering such signals in enough accuracy to prove useful for operational systems may be a further down the road.
An immediate application of the model presented in this work may be of quick magnitude estimation to be used for early warning systems. Such a system does not require additional data sources of assumptions on the model as the timing and location of the earthquake to be estimated are given by design of the task. 
A second application may incorporating such a magnitude predictor into an earthquake rate forecasting model such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}. In ETAS, the rate of events is determined by the product of a set of functions. \neri{formalize?} Origin time of events of the forecasted sequence are then sampled from the rate function, and assigned with a magnitude by sampling from a fitted GR distribution. By replacing the randomly assigned magnitudes of ETAS with MAGNET or a similar magnitude predictor one will improve the predictive power of rate prediction. In such a model, rate prediction will not be dominated by the abundant low magnitude events as is the case in ETAS, but rather by the high magnitude events which are of more importance for operational means. As we have shown - magnitudes predicted by MAGNET will estimate the true magnitudes in a higher likelihood.

\neri{I bet the reviewer will want us to talk about limitations and improvements to the model but maybe we can wait and see}
